Boutiques: a flexible framework for automated application integration in computing platforms

### 2016-10-28 paper.tex
Worked on configuration files and parallelization

\documentclass{article}
\usepackage{minted}
\usepackage{pdfcomment}

@@

\author{Tristan Glatard, Tristan Aumentado-Armstrong, Natacha Beck, \\
  Pierre Bellec, Remi Bernard, Sorina Camarasu-Pop, \\
  Fr\'ed\'eric Cervenansky, Samir Das, Rafael Ferreira da Silva, \\ 
  John Flavin, Pascal Girard, Krzysztof J. Gorgolewski, \\
  Nathaniel Kofalt, Pierre-Olivier Quirion, Pierre Rioux, \\
  Gunnar Schaeffer, Marc-\'Etienne Rousseau, Alan C. Evans}

@@

\begin{itemize}
\item JSON manifest
\item Link to container
\item Invocation schema
\end{itemize}

@@

The core component of the tool JSON manifest is a flexible
command-line description represented as a simple string complying to
the syntax of the \texttt{bash} Linux shell interpreter. The
command-line string may contain placeholders for input and output
values. The placeholders may be any string, and the command line may
actually encompass several commands, e.g., separated by semicolumns,
pipes (\texttt{|}) or ampersands (\texttt{\&}). Such a flexible
command-line specification is meant to facilitate minor adjustments on
the command-line, for instance input decompression and output
archival, avoiding the use of multiple ``shims'' to align mismatching
applications in a pipeline~\cite{hull2004treating}.

Here is an example of a typical command-line description:
\begin{verbatim}
exampleTool_1 [STRING_INPUT] [FILE_INPUT] [ENUM_INPUT] | \
exampleTool_2 [FLAG_INPUT] [NUMBER_INPUT] >> [LOG].txt
\end{verbatim}
The command line description contains six command-line keys for input
and output values. When the application is executed, such keys will be
replaced by numerical values and file names according to the user
input. No particular syntax is imposed for the command-line keys, they
just have to be unique. Flags will also be added wherever
appropriate. Note the use of the pipe operator to chain tools, and of
the \texttt{>>} bash operator that allows to redirect the standard
output in a particular file.

@@

\paragraph{General properties.} Inputs must have a name, a unique
identifier and a type. They may be optional, have a description, a
command-line key (placeholder), a flag and flag separator, and a
default value. Inputs may also be ordered lists.

\paragraph{Types.} Inputs may be of type \texttt{String},
\texttt{Number}, \texttt{Flag} or \texttt{File}. \texttt{File} may
also represent a directory. Types can be restricted to a specific set
of values (\texttt{String} and \texttt{Number} only), to a specific
range (\texttt{Number} only) to integers or booleans (\texttt{Number}
only).

\paragraph{Groups and dependencies.} Groups of inputs may be defined
from a identifier, name and list of input identifiers. Groups may be
used for presentation purposes in a graphical user interface and to
specify the following constraints among inputs: (1)
\texttt{mutually-exclusive}: only one input in the group may have a
value (2) \texttt{one-is-required}: at least one of the inputs in the
group must have a value. Dependencies among inputs may also be defined
regardless of a particular group: a specific input may (1) require a
list of other inputs to be active to be available and (2) define a
list of inputs that are disabled when it is active.

Listing~\ref{listing:input-example} shows the definition of an input
in the command line exemplified above. According to this definition
and assuming that the input value entered by the user is 0.3, the
string \texttt{[NUMBER\_INPUT]} will be replaced by \texttt{-n=0.3} on
the command line at execution time.
\begin{listing}
\begin{minted}[frame=single,
               framesep=3mm,
               linenos=false,
               xleftmargin=21pt,
               tabsize=4]{js}
{
    ``id'' : ``num_input'',
    ``name'' : ``A number input'',
    ``type'' : ``Number'',
    ``command-line-key'' : ``[NUMBER_INPUT]'',
    ``optional'' : true,
    ``command-line-flag'' : ``-n'',
    ``command-line-separator'' : ``='',
    ``minimum'' : 0,
    ``maximum'' : 1,
    ``exclusive-minimum'' : true,
    ``exclusive-maximum'' : false
  }
\end{minted}
\caption{Input example} 
\label{listing:input-example}
\end{listing}

@@

Application outputs are the files or directories that need to be
transferred and delivered to the user once the execution is
complete. While it may not be required to distinguish inputs and
outputs to build the application command line, this information is
needed by computing platforms to determine what needs to be saved from
the execution.

In Boutiques, output files are defined from a unique identifier, a
name and a path template that specifies the file or directory
name. Path templates may include input command-line keys when output
files are named after the input values. Input values may be stripped
from specific strings before being substituted in the path template,
e.g., file extensions. Output files may also have a description, a
command-line flag, a flag separator, and a command-line key in case
they appear on the command line. They may be optional in case the file
is not always produced by the application (for instance when it is
produced only when a particular flag is activated). They may also be
unordered lists; in this case, the path template must contain a
wildcard (\texttt{*}) standing for any string of characters (as in
\texttt{bash}) and defining the pattern used to match the output files
in the list.

@@

{
    "id" : "logfile",
    "name" : "Log file",
    "description" : "The output log file from the example tool",
    "path-template" : "log-[STRING_INPUT]",
    "command-line-key" : "[LOG]",
    "path-template-stripped-extensions" : [".txt", ".csv"],
    "optional" : false
}
\end{minted}
\caption{Output file example} 
\label{listing:output-example}
\end{listing}

@@

Input validation is an important motivation for common application
descriptions. The obvious solution in our context is to rely on an
application-specific JSON schema, called ``invocation schema'' to
specify the correct input values that an application
accepts. Platforms could leverage the invocation schema to validate
inputs automatically and without having to develop specific code,
using any JSON validator.

Invocation schemas, however, are complex JSON objects. Basically, they
must represent the properties described above in a formal way,
including dependencies between inputs. To illustrate such complexity,
listing~\ref{listing:invocation-schema-example} displays
\todo{Tristan}{...}. % must support dependent invocation ids to allow for workflow support.

It does not seem realistic to assume that
developers will easily write correct JSON schemas for their
applications. In addition, invocation schemas can actually be
generated automatically from the basic application description. Thus,
the invocation schema is an optional property of the Boutiques schema,
that platforms or developers may generate automatically using a core
tool provided by Boutiques.

A large number of applications actually rely on configuration files
rather than command-line options to define their input and output
parameters. Indeed, when the number of parameters increases,
command-line may become long and cumbersome while configuration files
allow for better structure and documentation.

Configuration files may be complex though, and specified in any
language.  As it does not seem reasonable to specify a particular
template for configuration files, the Boutiques specification allows
application developers to specify their own template, referring to
input and output command-line keys \todo{Tristan}{Then they are not
  command-line keys any more} to specify placeholders for input and
output values. Listing~\ref{listing:configuration-file-example} shows
an example. Configuration files also have a path template that defines
how they must be named and where they must be written. They may also
have a command line key and flag in case they need to be passed on the
command-line. 
\todo{Tristan}{See Github issue \#17}

\subsection{Parallelization support}

Instead, Boutiques allows to wrap complete workflows as simple
applications depending on particular workflow engine(s) installed in
container images. In this model, workflows are executed as any other
task on the infrastructure, which allows for basic data parallelism by
iterating the workflow on a set of independent files (for instance:
subjects in an analysis). Data parallelism, however, may not be
sufficient when the analysis of a particular file has to be
parallelized. To address this issue, a specific property of the
Boutiques schema specifies that an application may submit new
tasks. New tasks are created by the application by writing a JSON
object in the execution directory, complying to the invocation schema
described in Section~\ref{sec:invocation-schema}. The platform must
read this object, submit the corresponding task, and write back the
invocation identifier in the execution directory. The invocation
identifier may be used by the workflow to specify dependencies between
tasks in subsequent task submissions.

The adopted ``workflow encapsulation'' model is very simple but it
allows for a scalable and reliable execution of workflows expressed in
a variety of languages, as detailed in ~\cite{glatard2016fgcs}.

%\todo{rewrite this section to focus more on parallelism. Start by mentioning data parallelism.}

@@

Custom properties may be added to the Boutiques specification without
restriction. They may be useful to implement platform-specific
features but they should be used with care to avoid making tools
dependent on a particular platform. The following custom properties
have been used so far:
\begin{itemize}
\item \texttt{cbrain:can-submit-new-tasks}: a boolean, true if the tool may submit new tasks to the platform. \todo{Tristan}{related to the workflow support section}
\item \texttt{cbrain:inherits-from-class}: a string that defines the Ruby class that should be used as parent class for the
 tool in CBRAIN. Used to define a progress bar for PSOM tools in CBRAIN.
\item \texttt{vip:miccai-challenger-email} and
  \texttt{vip:miccai-challenge-team-name}: strings helping VIP
  categorize tools for the 2016 MICCAI challenges ``MSSEG'' and
  ``PETSEG''.
\end{itemize}

@@

The Boutiques validator checks conformance of JSON manifests to the
Boutiques schema using a basic JSON validator. It also performs an
additional set of checks that cannot be easily implemented in JSON
schema: command-line keys are unique among inputs, input and output
identifiers are unique, input and output command-line keys are all
included in the command-line, command-line keys are not contained
within each other (which would puzzle substitution), output path
templates are unique (to avoid results to overwrite each other),
inputs of type Flag have a command-line flag, are optional and are not
lists, number and list constraints are sensible (e.g. min is lower
than max), the default value of enum types is part of the enum, an
input cannot both require and disable another input, required inputs
cannot require or disable other parameters, group member identifiers
must correspond to existing inputs and cannot appear in different
groups, mutually exclusive groups cannot have members requiring other
members, and that one-is-required groups should never have required
members. The Boutiques validator is meant for application developers.

@@

@@

\section{References}
\bibliographystyle{plain}
\bibliography{biblio}

>>>>

\documentclass{article}
\usepackage{minted}
\usepackage{pdfcomment}
\usepackage[margin=1in]{geometry}

@@

\author{Tristan Glatard, Tristan Aumentado-Armstrong, Natacha Beck, Pierre Bellec,\\
        Remi Bernard, Sorina Camarasu-Pop, Fr\'ed\'eric Cervenansky, Samir Das, \\
        Rafael Ferreira da Silva,John Flavin, Pascal Girard, Krzysztof J. Gorgolewski, \\
        Nathaniel Kofalt, Pierre-Olivier Quirion, Pierre Rioux, Gunnar Schaeffer,\\
        Marc-\'Etienne Rousseau, Alan C. Evans}

@@

Define ``the platform'', i.e. the entity that will run applications.
\begin{itemize}
\item JSON manifest
\item Link to container
\item Invocation schema
\end{itemize}

@@

The core component of the JSON manifest is a flexible command-line
description represented as a simple string template complying to the
syntax of the \texttt{bash} Linux shell interpreter. The command line
template may contain placeholders for input and output values, called
value keys. The value keys may be any string, and the
command line may actually encompass several commands, e.g., separated
by semicolumns, pipes (\texttt{|}) or ampersands (\texttt{\&}). Such a
flexible command-line specification is meant to facilitate the
embedding of minor operations on the command line, for instance input
decompression and output archival, avoiding the use of multiple
``shims'' to align mismatching applications in a
pipeline~\cite{hull2004treating}.

Here is an example of a typical command-line template:
\begin{verbatim}
exampleTool_1 [STRING_INPUT] [FILE_INPUT] [ENUM_INPUT] | \
                              exampleTool_2 [FLAG_INPUT] [NUMBER_INPUT] >> [LOG].txt
\end{verbatim}
The template contains six value keys for input and output
values. When the application is executed, such keys will be replaced
by values and file names according to the user input. Flags will also
be added wherever appropriate, with customizable separators. No
particular syntax is imposed for the value keys, they just have
to be unique.  Note the use of the pipe (\texttt{|}) operator to chain
tools, and of the \texttt{>>} operator to redirect the standard output
to a file.

@@

\paragraph{General properties.} Inputs must have a name, a unique
identifier and a type. They may be optional, have a description, a
value key, a flag and flag separator, and a default
value. Inputs may also be ordered lists; in this case, value
keys are substituted by the space-separated list of input values.

\paragraph{Types.} Inputs may be of type \texttt{String},
\texttt{Number}, \texttt{Flag} or \texttt{File}. \texttt{File} may
also represent a directory. Types can be restricted to a specific set
of values (\texttt{String} and \texttt{Number} only), to a specific
range (\texttt{Number} only), to integers or to booleans (\texttt{Number}
only).

\paragraph{Groups and dependencies.} Groups of inputs may be defined
from a identifier, name and list of input identifiers. Groups may be
used to improve the presentation in a graphical user interface and to
specify the following constraints among inputs: (1)
\texttt{mutually-exclusive}: only one input in the group may have a
value; (2) \texttt{one-is-required}: at least one input in the group
must have a value. Dependencies among inputs may also be defined
regardless of a particular group: an input may (1) require a list of
inputs (2) disable a list of inputs.

Listing~\ref{listing:input-example} shows the definition of an input
in the command line exemplified above. According to this definition
and assuming that the input value entered by the user is 0.3, the
string \texttt{[NUMBER\_INPUT]} will be replaced by \texttt{-n=0.3} on
the command line at execution time.
\begin{listing}
\begin{minted}[frame=single,
               framesep=3mm,
               linenos=false,
               xleftmargin=21pt,
               tabsize=4]{js}
{
    "id" : "num_input",
    "name" : "A number input",
    "type" : "Number",
    "value-key" : "[NUMBER_INPUT]",
    "optional" : true,
    "command-line-flag" : "-n",
    "command-line-separator" : "=",
    "minimum" : 0,
    "maximum" : 1,
    "exclusive-minimum" : true,
    "exclusive-maximum" : false
  }
\end{minted}
\caption{Input example.} 
\label{listing:input-example}
\end{listing}

@@

Application outputs are the files or directories that need to be
transferred and delivered to the user once the execution is
complete. Although it may not be required to distinguish inputs from
outputs to build the application command line, this information is
needed by computing platforms to identify the files that must be
saved after the execution.

In Boutiques, output files are defined from a unique identifier, a
name and a path template that specifies the file or directory
name. Path templates may include input value keys in case
output files are named after the input values. Input values may be
stripped from specific strings, e.g., file extensions before being
substituted in the path template. Output files may also have a
description, a command-line flag, a flag separator, and a value
key in case they appear on the command line. They may be optional in
case the file is not always produced by the application, for instance
when it is produced only when a particular flag is activated. They
may also be (un-ordered) lists; in this case, the path template must
contain a wildcard (\texttt{*}) matching any string of characters
and defining the pattern used to match the
output files in the list.

@@

{
    "id" : "logfile",
    "name" : "Log file",
    "description" : "The output log file from the example tool",
    "path-template" : "log-[STRING_INPUT]",
    "value-key" : "[LOG]",
    "path-template-stripped-extensions" : [".txt", ".csv"],
    "optional" : false
}
\end{minted}
\caption{Output file example} 
\label{listing:output-example}
\end{listing}

A large number of applications rely on configuration files rather than
command-line options to define their input and output
parameters. Indeed, as the number of parameters increases, command
lines rapidly become long and cumbersome whereas configuration files
allow for better structure and documentation.

Configuration files may be complex though, and specified in any
language.  As it does not seem reasonable to adopt a particular
template for configuration files, the Boutiques specification allows
application developers to specify their own template, referring to
input and output value keys. Configuration files are specific types of
output files that have a
 file template defined. As output files, 
they must have a path template that defines
how they must be named and where they must be written. They may also
have a value key and a flag in case they need to be passed on the
command line.

Listing~\ref{listing:configuration-file-example} shows an example.
\begin{listing}
\begin{minted}[frame=single,
               framesep=3mm,
               linenos=false,
               xleftmargin=21pt,
               tabsize=4]{js}
               {
                   "id": "config_file",
                   "name": "Configuration file",
                   "type": "Configuration File",
                   "value-key": "[CONFIG_FILE]",
                   "path-template": "config.txt",
                   "file-template": [
                   "# This input is hard-coded",
                   "stringInput=foo",
                   "fileInput=[FILE-INPUT]",
                   "# And here is the result",
                   "fileOutput=[OUTPUT-FILE]",
                   ""
                   ]
                 }
\end{minted}
\caption{Configuration file example. The file template is defined as
  an array of strings to allow for multi-line strings in JSON.}
\label{listing:configuration-file-example}
\end{listing}

@@

Rigorous input validation is an important motivation for
Boutiques. The obvious solution in our context is to rely on an
application-specific JSON schema, called ``invocation schema'', to
specify the correct input values that an application
accepts. Platforms could rely on the invocation schema to validate
inputs automatically using any JSON validator, without having to
develop specific code.

Invocation schemas, however, are complex JSON objects. Basically, they
must represent the properties described above in a formal way,
including dependencies between
inputs. Listing~\ref{listing:invocation-schema-example} shows an
example of how dependencies between mutually exclusive parameters are
defined in the invocation schema.

The invocation schema also has a \texttt{task-dependencies} array that
specifies that the new invocation to start only when all the
invocations in this array have completed. This property is used in the
parallelization model described hereafter.

\begin{listing}
\begin{minted}[frame=single,
               framesep=3mm,
               linenos=false,
               xleftmargin=21pt,
               tabsize=4]{js}
"dependencies" : {
      "num_input" : {
         "properties" : {
            "str_input" : {
               "not" : {}
            }
         }
      },
      "str_input" : {
         "properties" : {
            "num_input" : {
               "not" : {}
            }
         }
      }
   }
\end{minted}
\caption{Excerpt from invocation schema showing dependencies between
  two mutually exclusive parameters \texttt{num\_input} and
  \texttt{str\_input}.}
\label{listing:invocation-schema-example}
\end{listing}

It does not seem realistic to assume that developers will easily write
correct JSON schemas for their applications. Instead, invocation
schemas should be generated automatically using a core Boutiques
tool. Accordingly, invocation schema is an optional property of the
Boutiques manifest, that platforms or developers may generate using
the appropriate tool.

\subsection{Parallelization support}
\label{sec:parallelization}
Boutiques support parallelism by allowing applications to submit new
tasks through the following protocol:
\begin{enumerate}
\item Applications write a JSON object complying to the invocation schema in the directory where there were launched.
\item The platform reads the JSON object, creates the corresponding task, and writes back the task id in the same directory.
% The platform could also support status requests.
\end{enumerate}
Several invocations can be submitted at once, possibly using the
\texttt{task-dependencies} array to specify dependencies among
them. This mechanism is implemented in the Boutiques schema using a
single property, \texttt{can-submit-tasks}.

Although very simple, this model allows wrapping complete
workflows. Workflows are wrapped as any other application, and they
can be executed using any engine such as Nipype, PSOM and others,
provided that such engines are installed in the container image
referred in the application manifest.

The adopted ``task encapsulation'' model is very simple but it
allows for a scalable and reliable execution of workflows expressed in
a variety of languages, as detailed in ~\cite{glatard2016fgcs}.

@@

Custom properties may be added to the Boutiques specification without
restriction. They may be useful to implement platform-specific
features but they should be used with care to avoid making tools
dependent on a particular platform. The following custom properties
have been used so far:
\begin{itemize}
\item \texttt{cbrain:inherits-from-class}: a string that defines the Ruby class that should be used as parent class for the
 tool in CBRAIN. Used to define a progress bar for PSOM tools in CBRAIN.
\item \texttt{vip:miccai-challenger-email} and
  \texttt{vip:miccai-challenge-team-name}: strings helping VIP
  categorize tools for the 2016 MICCAI challenges ``MSSEG'' and
  ``PETSEG''.
\end{itemize}

@@

The Boutiques validator checks conformance of JSON manifests to the
Boutiques schema using a basic JSON validator. It also performs an
additional set of checks that cannot be easily implemented in JSON
schema: value keys are unique among inputs, input and output
identifiers are unique, input and output value keys are all
included in the command line, value keys are not contained
within each other (which would puzzle substitution), output path
templates are unique (to avoid results to overwrite each other),
inputs of type Flag have a command-line flag, are optional and are not
lists, number and list constraints are sensible (e.g. min is lower
than max), the default value of enum types is part of the enum, an
input cannot both require and disable another input, required inputs
cannot require or disable other parameters, group member identifiers
must correspond to existing inputs and cannot appear in different
groups, mutually exclusive groups cannot have members requiring other
members, and that one-is-required groups should never have required
members. The Boutiques validator is meant for application developers.

@@

% How to build modular container images for workflows.

@@

\bibliographystyle{plain}
\bibliography{biblio}
